A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-26 sections/Related.tex
Introduction

\paragraph{Impredicative Higher-Ranked Polymorphism Type Inference Algorithms}
Impredicative System F allows instantiation with polymorphic types,
but unfortunately its subtyping system is already undecidable~\cite{tiuryn1996subtyping}.
Works on partial impredicative type-inference algorithms\cite{le2003ml,leijen2008hmf,vytiniotis2008fph}
navigate a variety of design tradeoffs for a decidable algorithm.
As a result, such algorithms tend to be more complicated, and thus less adopted in practise.
\jimmy{Guarded Impredicative}
A recent work proposed Guarded Impredicative Polymorphism~\cite{},
as an improvement on GHC's type inference algorithm with impredicative instantiation.
They make use of local informations in $n$-ary applications to
infer polymorphic instantiations with a relatively simple specification and unification algorithm.
Although not all impredicative instantiations can be handled well,
their algorithm is already quite useful in practice.

>>>>

\paragraph{Impredicative Higher-Ranked Polymorphism Type Inference Algorithms}
Impredicative System F allows instantiation with polymorphic types,
but unfortunately its subtyping system is already undecidable~\cite{tiuryn1996subtyping}.
Works on partial impredicative type-inference algorithms\cite{le2003ml,leijen2008hmf,vytiniotis2008fph}
navigate a variety of design tradeoffs for a decidable algorithm.
As a result, such algorithms tend to be more complicated, and thus less adopted in practise.
\jimmy{Guarded Impredicative}
A recent work proposed Guarded Impredicative Polymorphism~\cite{Serrano2018},
as an improvement on GHC's type inference algorithm with impredicative instantiation.
They make use of local informations in $n$-ary applications to
infer polymorphic instantiations with a relatively simple specification and unification algorithm.
Although not all impredicative instantiations can be handled well,
their algorithm is already quite useful in practice.

### 2019-02-26 sections/introduction.tex
Introduction

As functional languages evolved the need for more expressive
power has motivated language designers to look beyond Hindley-Milner.
In particular one popular direction is to allow 
\emph{higher-ranked polymorphism} where polymorphic types can
occur anywhere in a type signature.  
An important challenge is that full type inference for higher-ranked
polymorphism is know to be undecidable~\cite{wells1999typability}. Therefore some type
annotations are necessary to guide type inference. In response to this 
challenge several decidable type systems requiring some annotations 
have been proposed~\cite{,dunfield2013complete,jones2007practical}
\bruno{works on impredicative type inference are still missing}.
Two closely related type systems that 
support \emph{predicative} higher-ranked type inference were proposed 
by Peyton Jones et al.~\cite{jones2007practical} and Dunfield and
Krishnaswami~\cite{dunfield2013complete} (henceforth denoted as DK). 
These type systems are
popular among language designers and their ideas have been adopted by
several modern functional languages, including Haskell, PureScript~\cite{} and
Unison~\bruno{Add references. Here is the one for unison\url{http://unisonweb.org}} among others.
In those type systems
type annotations are required for polymorphic arguments of functions,
but other type annotations can be omitted. A canonical example (here written in Haskell) is:
\begin{verbatim}
    hpoly = (\f :: forall a. a -> a) -> (f 1, f 'c')
\end{verbatim}
The function \verb|hpoly| cannot be
type-checked in Hindley-Milner.  The type of \verb|hpoly| is the rank-2 type:
\verb|(forall a. a -> a) -> (Int, Char)|. Notably (and unlike
Hindley-Milner) the lambda argument \verb|f| requires a
\emph{polymorphic} type annotation.
This annotation is needed because the single universal quantifier
does not appear at the top-level. Instead it is used to quantify a
type variable \verb|a| used in the first argument of the
function. 
Despite these additional annotations, Peyton Jones et al. and DK's
type inference algorithms preserve many of the desirable properties 
of Hindley-Milner. For example the applications of \verb|f| implicitly 
instantiate the polymorphic type arguments of \verb|f|.

Although type inference is important in practice and receives a lot of
attention in
academic research, there is little work on mechanically formalizing
such advanced forms of type inference in theorem provers.
The remarkable exception is some work done on the formalization of 
certain parts of Hindley-Milner type inference~\cite{naraschewski1999type,
dubois2000proving,dubois1999certification,urban2008nominal,
garrigue2015certified}. However
there is still no formalization of the higher-ranked type systems
that are employed by modern languages like Haskell.
This is at
odds with the current trend of mechanical formalizations in
programming language research. In particular both the POPLMark
challenge~\cite{aydemir2005mechanized} and
CompCert ~\cite{leroy2012compcert} have significantly promoted
the use of theorem provers to model various aspects of programming
languages. Today papers in various programming language venues routinely
use theorem provers to mechanically formalize: \emph{dynamic and
  static semantics} and their correctness properties~\cite{aydemir2008engineering},
\emph{compiler correctness}~\cite{leroy2012compcert}, \emph{correctness of
  optimizations}~\cite{Bertot04}, \emph{program analisys}~\cite{Chang2006}
or proofs involving \emph{logical relations}~\cite{abel2018}. The
main argument for mechanical formalizations is a simple one. Proofs
for programming languages tend to be \emph{long}, \emph{tedious} and
\emph{error-prone}. In such proofs it is very easy to make mistakes
that may invalidate the whole development. Furthermore, readers and
reviewers often do not have time to look at the proofs carefully to
check their correctness. Therefore errors can go unnoticed for a
long time.  Mechanical formalizations provide, in principle, a natural
solution for these problems. Theorem provers can automatically check and
validate the proofs, which removes the burden from checking from both
the person doing the proofs as well as readers or reviewers.

@@

Due to those two challenges, our work takes a different approach by refining and
extending the idea of \emph{worklists judgments}~\cite{itp2018},
proposed recently to mechanically formalize an algorithm for
\emph{polymorphic subtyping}. A key innovation in our work is how
to adapt the idea of worklist judgments to
\emph{inference judgements}, which are not needed for polymorphic subtyping.
In our work we show how worklist
judgements can be extended to deal with the inference judgements that
are necessary for type-inference.  The idea is to use a continuation
passing style to enable the transfer of inferred information across
judgements. A further refinement to the idea of worklist judgements is
the \emph{unification between ordered
  contexts~\cite{gundry2010type,dunfield2013complete} and worklists}.  This
enables precise scope tracking of free variables in
judgements. Furthermore it avoids the duplication of context
information across judgments in worklists that occurs in other
techniques~\cite{Reed2009,Abel2011higher}.
Despite the use of a different algorithm we prove the
same results as DK, although with significantly different proofs and
proof techniques. The calculus and its metatheory
have been fully formalized in the Abella theorem prover~\cite{AbellaDesc}.

@@

\item {\bf Unification of judgment worklists and contexts:} Another technical contribution is the unification
  between ordered contexts and judgement worklists. This enables precise scope tracking
  of variables in judgements, and avoids the duplication of context information across
  judgments in worklists.

>>>>

As functional languages evolved the need for more expressive
power has motivated language designers to look beyond Hindley-Milner.
In particular one popular direction is to allow 
\emph{higher-ranked polymorphism} where polymorphic types can
occur anywhere in a type signature.  
An important challenge is that full type inference for higher-ranked
polymorphism is know to be undecidable~\cite{wells1999typability}. Therefore some type
annotations are necessary to guide type inference. In response to this 
challenge several decidable type systems requiring some annotations 
have been proposed~\cite{,dunfield2013complete,jones2007practical,Serrano2018,le2003ml,leijen2008hmf,vytiniotis2008fph}.
Two closely related type systems that 
support \emph{predicative} higher-ranked type inference were proposed 
by Peyton Jones et al.~\cite{jones2007practical} and Dunfield and
Krishnaswami~\cite{dunfield2013complete} (henceforth denoted as DK). 
These type systems are
popular among language designers and their ideas have been adopted by
several modern functional languages, including Haskell, PureScript~\cite{PureScript} and
Unison~\cite{Unison} among others.
In those type systems
type annotations are required for polymorphic arguments of functions,
but other type annotations can be omitted. A canonical example (here written in Haskell) is:
\begin{verbatim}
    hpoly = (\f :: forall a. a -> a) -> (f 1, f 'c')
\end{verbatim}
The function \verb|hpoly| cannot be
type-checked in Hindley-Milner.  The type of \verb|hpoly| is the rank-2 type:
\verb|(forall a. a -> a) -> (Int, Char)|. Notably (and unlike
Hindley-Milner) the lambda argument \verb|f| requires a
\emph{polymorphic} type annotation.
This annotation is needed because the single universal quantifier
does not appear at the top-level. Instead it is used to quantify a
type variable \verb|a| used in the first argument of the
function. 
Despite these additional annotations, Peyton Jones et al. and DK's
type inference algorithms preserve many of the desirable properties 
of Hindley-Milner. For example the applications of \verb|f| implicitly 
instantiate the polymorphic type arguments of \verb|f|.

Although type inference is important in practice and receives a lot of
attention in
academic research, there is little work on mechanically formalizing
such advanced forms of type inference in theorem provers.
The remarkable exception is work done on the formalization of 
certain parts of Hindley-Milner type inference~\cite{naraschewski1999type,
dubois2000proving,dubois1999certification,urban2008nominal,
garrigue2015certified}. However
there is still no formalization of the higher-ranked type systems
that are employed by modern languages like Haskell.
This is at
odds with the current trend of mechanical formalizations in
programming language research. In particular both the POPLMark
challenge~\cite{aydemir2005mechanized} and
CompCert ~\cite{leroy2012compcert} have significantly promoted
the use of theorem provers to model various aspects of programming
languages. Today papers in various programming language venues routinely
use theorem provers to mechanically formalize: \emph{dynamic and
  static semantics} and their correctness properties~\cite{aydemir2008engineering},
\emph{compiler correctness}~\cite{leroy2012compcert}, \emph{correctness of
  optimizations}~\cite{Bertot04}, \emph{program analisys}~\cite{Chang2006}
or proofs involving \emph{logical relations}~\cite{abel2018}. The
main argument for mechanical formalizations is a simple one. Proofs
for programming languages tend to be \emph{long}, \emph{tedious} and
\emph{error-prone}. In such proofs it is very easy to make mistakes
that may invalidate the whole development. Furthermore, readers and
reviewers often do not have time to look at the proofs carefully to
check their correctness. Therefore errors can go unnoticed for a
long time.  Mechanical formalizations provide, in principle, a natural
solution for these problems. Theorem provers can automatically check and
validate the proofs, which removes the burden from checking from both
the person doing the proofs as well as readers or reviewers.

@@

Due to those two challenges, our work takes a different approach by refining and
extending the idea of \emph{worklists judgments}~\cite{itp2018},
proposed recently to mechanically formalize an algorithm for
\emph{polymorphic subtyping}~\cite{odersky1996putting}. A key innovation in our work is how
to adapt the idea of worklist judgments to
\emph{inference judgements}, which are not needed for polymorphic
subtyping, but are necessary for type-inference.  The idea is to use a \emph{continuation
passing style} to enable the transfer of inferred information across
judgements. A further refinement to the idea of worklist judgements is
the \emph{unification between ordered
  contexts~\cite{gundry2010type,dunfield2013complete} and worklists}.  This
enables precise scope tracking of free variables in
judgements. Furthermore it avoids the duplication of context
information across judgments in worklists that occurs in other
techniques~\cite{Reed2009,Abel2011higher}.
Despite the use of a different algorithm we prove the
same results as DK, although with significantly different proofs and
proof techniques. The calculus and its metatheory
have been fully formalized in the Abella theorem prover~\cite{AbellaDesc}.

@@

\item {\bf Unification of worklists and contexts:} Another technical contribution is the unification
  between ordered contexts and worklists. This enables precise scope tracking
  of variables in judgements, and avoids the duplication of context information across
  judgments in worklists.
